---
title: "Functional Programming: Linear Model with RBF Feature Transform"
output: pdf_document
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
- \usepackage{bm}
- \usepackage{mathtools}
---
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
\newcommand{\norm}[1]{||#1||}
\newcommand{\vx}{\vect{x}}
\newcommand{\vphi}{\vect{\phi}}
\newcommand{\vX}{\vect{X}}
\newcommand{\vxi}{\vect{\xi}}
\newcommand{\delement}[2]{\vx_{#1}^\top\vx_{#1} - 2\vx_{#1}^\top \vx_{#2} + \vx_{#2}^\top\vx_{#2}}>
\newcommand{\hvx}{\widehat{\vx}}
\newcommand{\nrm}[1]{\parallel #1 \parallel}
\newcommand{\mydot}[2]{\vx_{#1}^\top\vx_{#2}}
\newcommand{\mydothat}[2]{\hvx_{#1}^\top\vx_{#2}}
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Mathematical Set Up
Suppose we have a training matrix $X$ with $n$ observations and $d$ features
$$
X = \begin{pmatrix}
x_{11} & x_{12} & \ldots & x_{1d}\\
x_{21} & x_{22} & \ldots & x_{2d} \\
\vdots & \vdots & \ddots & \vdots \\
x_{n1} & x_{n2} & \ldots & x_{nd}
\end{pmatrix}
=\begin{pmatrix}
{\bf{x}}_1^\top \\
{\bf{x}}_2^\top \\
\vdots \\
{\bf{x}}_n^\top
\end{pmatrix}
\in\mathbb{R}^{n\times d}
$$
and that we want to do a **feature transform** of this data using the Radial Basis Function. To do this, we 

* choose $b$ rows of $X$ and we call them **centroids**
$$
{\bf{x}}^{(1)}, \ldots, {\bf{x}}^{(b)}
$$
* calculate using some heuristic a **bandwidth** parameter $\sigma^2$

And then, for every centroid we define a radial basis function as follows
$$
\phi^{(i)}({\bf{x}}):=\exp\left(- \frac{\parallel{\bf{x}} - {\bf{x}}^{(i)}\parallel^2}{\sigma^2}\right) \qquad \forall i\in\{1, \ldots, b\} \quad \text{for } {\bf{x}}\in\mathbb{R}^{d}
$$
We can therefore obtain a transformed data matrix as
$$
\Phi:=\begin{pmatrix}
1 & \phi^{(1)}({\bf{x}}_1) & \phi^{(2)}({\bf{x}}_1) & \cdots & \phi^{(b)}({\bf{x}}_1) \\
1 & \phi^{(1)}({\bf{x}}_2) & \phi^{(2)}({\bf{x}}_2) & \cdots & \phi^{(b)}({\bf{x}}_2) \\
\vdots & \vdots & \vdots & \ddots & \vdots \\
1 & \phi^{(1)}({\bf{x}}_n) & \phi^{(2)}({\bf{x}}_n) & \cdots & \phi^{(b)}({\bf{x}}_n)
\end{pmatrix} \in\mathbb{R}^{n\times (b+1)}
$$
Then we fit a **regularized** linear model so that **optimal parameters** are given by
$$
{\bf{w}}:=(\Phi^\top\Phi + \lambda I_n)^{-1}\Phi^\top{\bf{y}}
$$
These parameters are usually found by solving the associated system of linear equations for ${\bf{w}}$ 
$$
(\Phi^\top\Phi + \lambda I_n) {\bf{w}} = \Phi^\top {\bf{y}}
$$
This will be one of the strategies that we'll see below. However, we are required to apply regularization with a very small regularization parameter $\lambda$. This hints at the fact that maybe we can simply set $\lambda=0$ and then find ${\bf{w}}$ simply by finding the pseudoinverse of $\Phi$
$$
(\Phi^\top\Phi)^{-1}\Phi^\top
$$
and then multiply this by ${\bf{y}}$ on the left. This approach looks more stable and can be implemented by setting `pseudoinverse=TRUE` in the function `make_predictor` defined in this document.

## Parameters
In the following code chunk you can set up $n$ the number of training observations, $d$ the dimensionality of the input (i.e. the number of features), $\lambda$ the regularization coefficient, $m$ the number of testing observations, and $b$ dimensionality of the data after the feature transform, which in this case corresponds to the number of centroids.
```{r}
n <-  1000
d <-  2
lambda <-  0.00001
m <-  200
b <-  50
```

## Training and Testing Sets
We work with synthetic data sets. For simplicity, we will generate $X$ uniformly in an interval (which in this case is $[-4, 4]$) and then $y$ will be taken to be a multivariate normal of each row of $X$.
Note that we create a **factory of functions** to generate multivariate normal
closures, yet the same behavior would be achieved writing up a simple function.
The factory of functions is defined in order to work with the functional 
programming paradigm.
```{r}
# Explanatory variable is uniformly distributed between -4 and 4, then reshaped
X <-  matrix(runif(n*d, min=-4, max=4), nrow=n, ncol=d)
# Factory of multivariate normal distributions
normal_factory <- function(mean_vector, cov_matrix){
  d <- nrow(cov_matrix)
  normal_pdf <- function(x){
    exponent <- -mahalanobis(x, center=mean_vector, cov=cov_matrix) / 2
    return(
      (2*pi)^(-d/2) * det(cov_matrix)^(-1/2) * exp(exponent)
    )
  }
}
# Closure will be our data-generating process for training and test response.
target_normal <- normal_factory(rnorm(d), diag(d))
y <- matrix(apply(X, 1, target_normal))
```
Similarly, create test data.
```{r}
Xtest <- matrix(runif(m*d, min=-4, max=4), nrow=m, ncol=d)
ytest <- matrix(apply(Xtest, 1, target_normal))
```
## Feature Transform Implementation
Define a function that, given a training data matrix $X$, finds the distance matrix containing all the pairwise distances squared, and then return the median of such distances. Essentially, it returns the badwidth squared $\sigma^2$.
```{r}
compute_sigmasq <- function(X){
  # Compute distance matrix expanding the norm. Return its median
  Xcross <- tcrossprod(X)
  Xnorms <- matrix(diag(Xcross), nrow(X), nrow(X), byrow=TRUE)
  return(median(Xnorms - 2*Xcross + t(Xnorms)))
}
```
Define a function that gets $b$ random samples from the rows of an input matrix $X$, these will be the centroids.
```{r}
get_centroids <- function(X, n_centroids){
  # Find indeces of centroids
  idx <- sample(1:nrow(X), n_centroids)
  return(X[idx, ])
}
```
Construct a factory of Radial Basis Functions. The factory constructs an RBF, given a centroid and a $\sigma^2$.
```{r}
rbf <- function(centroid, sigmasq){
  rbfdot <- function(x){
    return(
      exp(-sum((x - centroid)^2) / sigmasq)
    )
  }
  return(rbfdot)
}
```
Finally, we can put all of these functions together to compute $\Phi$ and eventually make predictions. 
```{r}
compute_phiX <- function(X, centroids, sigmasq){
  # Create a list of rbfdot functions and apply each of them to every row of X
  phiX <- mapply(
    function(f) apply(X, 1, f), 
    apply(centroids, 1, rbf, sigmasq=sigmasq)
  )
  # Reshape phiX correctly
  phiX <- matrix(phiX, nrow(X), nrow(centroids))
  # Recall we need the first column to contain 1s for the bias
  return(cbind(1, phiX))  # this is a (n, d+1) matrix
}
```
## Prediction
We're now ready to define a factory of functions that returns prediction functions.
```{r}
library(corpcor)
make_predictor <- function(X, y, lambda, n_centroids, pseudoinverse=FALSE){
  # Randomly sample centoids
  centroids <- get_centroids(X, n_centroids)
  sigmasq <- compute_sigmasq(X)
  # Get transformed data matrix
  phiX <- compute_phiX(X, centroids, sigmasq)
  # Find optimal parameters
  if (pseudoinverse){
    # This method works best cause it does automatic regularization with SVD
    w <- pseudoinverse(phiX) %*% y
  } else {
    w <- solve(t(phiX)%*%phiX + lambda*diag(n_centroids+1), t(phiX) %*% y)
  }
  # Construct predictor closure for a new batch of testing data Xtest
  predictor <- function(Xtest){
    # Need to transform test data into a phi matrix
    phi_Xtest <- compute_phiX(Xtest, centroids, sigmasq)
    return(phi_Xtest %*% w)
  }
  return(predictor)
}
```
We can test its effectiveness now. Notice that in general we might need a lot of centroids to make this work.
```{r}
# Construct prediction functions. One with regularization, other with pseudoinv
predict <- make_predictor(X, y, lambda, n_centroids = b)
pseudo_predict <- make_predictor(X, y, lambda, n_centroids = b, pseudoinverse = TRUE)
# Get predictions
yhat <- predict(Xtest)
yhat_pseudo <- pseudo_predict(Xtest)
```
Predictions can be assessed by plotting predicted values against the true test 
values. 
```{r}
library(ggplot2)
# Put data into a dataframe for ggplot2
df <- data.frame(real=ytest, pred=yhat, pseudo_pred=yhat_pseudo)
ggplot(data=df) + 
  geom_point(aes(x=real, y=pred), color="darkblue") + 
  geom_abline(intercept=0, slope=1) + 
  ggtitle(paste("Predictions vs Actual with b =", b)) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size=15, face="bold"),
    axis.title.y = element_text(angle=0, vjust=0.5)
  ) +
  xlab(expression(y)) + 
  ylab(expression(hat(y))) + 
  coord_fixed()
```

Similarly, we can see the performance of our prediction using the pseudoinverse
```{r}
ggplot(data=df) + 
  geom_point(aes(x=real, y=pseudo_pred), color="darkblue") + 
  geom_abline(intercept=0, slope=1) + 
  ggtitle(paste("Pseudo-Predictions vs Actual with b =", b)) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.title = element_text(size=15, face="bold"),
    axis.title.y = element_text(angle=0, vjust=0.5)
  ) +
  xlab(expression(y)) + 
  ylab(expression(hat(y))) + 
  coord_fixed()
```

## Technical Note on Calculating the Bandwidth
To vectorize this operation, we aim to create a matrix containing all the squared norms of the difference between the vectors
$$
D = 
\begin{pmatrix}
  \nrm{\vx_1 - \vx_1}^2 & \nrm{\vx_1 - \vx_2}^2 & \cdots & \nrm{\vx_1 - \vx_n}^2 \\
  \nrm{\vx_2 - \vx_1}^2 & \nrm{\vx_2 - \vx_2}^2 & \cdots & \nrm{\vx_2 - \vx_n}^2 \\
  \vdots             & \vdots             & \ddots & \vdots \\
  \nrm{\vx_n - \vx_1}^2 & \nrm{\vx_n - \vx_2}^2 & \cdots & \nrm{\vx_n - \vx_n}^2
\end{pmatrix}
$$
Notice that we can rewrite the norm of the difference between two vectors $\vx_i$ and $\vx_j$ as follows
$$
\nrm{\vx_i - \vx_j}^2 = (\vx_i - \vx_j)^\top (\vx_i - \vx_j) = \vx_i^\top\vx_i - 2\vx_i^\top \vx_j + \vx_j^\top\vx_j
$$
Broadcasting this, we can rewrite the matrix $D$ as
$$
D =
\begin{pmatrix}
  \delement{1}{1} & \quad\delement{1}{2} & \quad\cdots & \quad\delement{1}{n}\\
  \delement{2}{1} & \quad\delement{2}{2} & \quad\cdots & \quad\delement{2}{n}\\
  \vdots          & \quad\vdots          & \quad\ddots & \quad\vdots \\
  \delement{n}{1} & \quad\delement{n}{2} & \quad\cdots & \quad\delement{n}{n}
\end{pmatrix}
$$
This can now be split into three matrices, where we can notice that the third matrix is nothing but the transpose of the first.
$$
D =
\begin{pmatrix}
\vx_1^\top\vx_1 & \vx_1^\top\vx_1 & \cdots & \vx_1^\top\vx_1\\
\vx_2^\top\vx_2 & \vx_2^\top\vx_2 & \cdots & \vx_2^\top\vx_2 \\
\vdots          & \vdots          & \ddots & \vdots\\
\vx_n^\top\vx_n & \vx_n^\top\vx_n & \cdots & \vx_n^\top\vx_n
\end{pmatrix}
-2
\begin{pmatrix}
  \mydot{1}{1} & \mydot{1}{2} & \cdots & \mydot{1}{n} \\
  \mydot{2}{1} & \mydot{2}{2} & \cdots & \mydot{2}{n} \\
  \vdots     & \vdots     & \ddots & \vdots \\
  \mydot{n}{1} & \mydot{n}{2} & \cdots & \mydot{n}{n}
\end{pmatrix}
+
\begin{pmatrix}
\mydot{1}{1} & \mydot{2}{2} & \cdots & \mydot{n}{n} \\
\mydot{1}{1} & \mydot{2}{2} & \cdots & \mydot{n}{n} \\
\vdots     & \vdots     & \ddots & \vdots \\
\mydot{1}{1} & \mydot{2}{2} & \cdots & \mydot{n}{n} 
\end{pmatrix}
$$

The key is not to notice that we can obtain the middle matrix from $X$ with a simple operation:
$$
XX^\top = 
\begin{pmatrix}
  \vx_1^\top \\
  \vx_2^\top \\
  \vdots \\
  \vx_n^\top
\end{pmatrix}
\begin{pmatrix}
\vx_1 & \vx_2 & \cdots & \vx_n 
\end{pmatrix}
=
\begin{pmatrix}
  \mydot{1}{1} & \mydot{1}{2} & \cdots & \mydot{1}{n} \\
  \mydot{2}{1} & \mydot{2}{2} & \cdots & \mydot{2}{n} \\
  \vdots     & \vdots     & \ddots & \vdots \\
  \mydot{n}{1} & \mydot{n}{2} & \cdots & \mydot{n}{n}
\end{pmatrix}
$$

and then the first matrix can be obtained as follows
$$
\begin{pmatrix}
  1 \\
  1 \\
  \vdots \\
  1
\end{pmatrix}_{n\times 1}
\begin{pmatrix}
  \mydot{1}{1} & \mydot{2}{2} & \cdots & \mydot{n}{n}
\end{pmatrix}_{1\times n}
= 
\begin{pmatrix}
\vx_1^\top\vx_1 & \vx_1^\top\vx_1 & \cdots & \vx_1^\top\vx_1\\
\vx_2^\top\vx_2 & \vx_2^\top\vx_2 & \cdots & \vx_2^\top\vx_2 \\
\vdots          & \vdots          & \ddots & \vdots\\
\vx_n^\top\vx_n & \vx_n^\top\vx_n & \cdots & \vx_n^\top\vx_n
\end{pmatrix}
$$
where
$$
\text{diag}\left[
\begin{pmatrix}
  \mydot{1}{1} & \mydot{1}{2} & \cdots & \mydot{1}{n} \\
  \mydot{2}{1} & \mydot{2}{2} & \cdots & \mydot{2}{n} \\
  \vdots     & \vdots     & \ddots & \vdots \\
  \mydot{n}{1} & \mydot{n}{2} & \cdots & \mydot{n}{n}
\end{pmatrix}
\right]
=
\begin{pmatrix}
  \mydot{1}{1} & \mydot{2}{2} & \cdots & \mydot{n}{n}
\end{pmatrix}_{1\times n}
$$








