---
title: "examplegp"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Import libraries
```{r}
library(MASS)
library(reshape2)
library(ggplot2)
```

Define points for the function values and define a bandwidth `l`
```{r}
x_predict <- seq(-5,5, len=50)
l <- 1.0
```
Define a squared exponential (`SE`) function and a covariance function `cov`
```{r}
SE <- function(Xi, Xj, l){
  return(exp(-0.5 * (Xi - Xj)^2 / l^2))
}
cov <- function(X, Y){
  return(outer(X, Y, SE, l))
}
```
Now calculate the `K` matrix, basically this will be the `COV` matrix here. 
```{r}
COV <- cov(x_predict, x_predict)
```
Sample some function
```{r}
values <- mvrnorm(3, rep(0, length=length(x_predict)), COV)
```
Melt to put in long form to be able to plot
```{r}
dat <- data.frame(x=x_predict, t(values))
dat <- melt(dat, id="x")
head(dat)
```
Finally plot it
```{r}
ggplot(data=dat, aes(x=x, y=value)) +
  geom_rect(xmin=-Inf, xmax=Inf, ymin=-2, ymax=2, fill="grey80") + 
  geom_line(aes(group=variable)) + 
  scale_y_continuous(lim=c(-3, 3), name="output, f(x)") + 
  xlab("input, x") + 
  theme_bw()
```

## Noise-free observations
Here is some observed data
```{r}
obs <- data.frame(x=c(-4, -3, -1, 0, 2),
                  y=c(-2,  0, 1, 2, -1))
```
Now we calculate
$$
cov(X_0, X_0)^{-1}
$$
```{r}
cov_xx_inv <- solve(cov(obs$x, obs$x))
```
Now find the predictive mean
$$
Ef = cov(X_p, X_0)cov(X_0, X_0)^{-1}y
$$

```{r}
Ef <- cov(x_predict, obs$x) %*% (cov_xx_inv %*% obs$y)
```

Now compute the variance covariance matrix
$$
C = cov(X_p, X_p) - cov(X_p, X_0)cov(X_0, X_0)^{-1}cov(X_0, X_p)
$$
```{r}
Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x) %*% cov_xx_inv %*% cov(obs$x, x_predict)
```

Get $3$ samples from the posterior
```{r}
values <- mvrnorm(3, Ef, Cf)
```


Plot the solutions
```{r}
# Create a dataframe with x values and the new samples
dat <- data.frame(x=x_predict, t(values))
dat <- melt(dat, id="x")
# dataframe with the mean
dfmean <- data.frame(x=x_predict, y=Ef, ymin=Ef-2*sqrt(diag(Cf)), ymax=Ef+2*sqrt(diag(Cf)))
# Plot the figure
ggplot() + 
  geom_ribbon(data=dfmean, aes(x=x_predict, ymin=ymin, ymax=ymax), fill="grey80") +
  geom_line(data=dat, aes(x=x, y=value, color=variable)) + 
  geom_line(data=dfmean, aes(x=x_predict, y=Ef), size=1) + 
  geom_point(data=obs, aes(x=x, y=y), color='red') + 
  xlab("input, x") + 
  ylab("output, f(x)")
```


Now we work with additive noise instead. We choose a sigma that represents the variance of the zero-mean noise added to each observation.

```{r}
sigma.n <- 0.25
cov_xx_inv <- solve(cov(obs$x, obs$x) + sigma.n^2 * diag(1, length(obs$x)))
Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% obs$y
Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x) %*% cov_xx_inv %*% cov(obs$x, x_predict)
```

Get $3$ random samples now

```{r}
values <- mvrnorm(3, Ef, Cf)
```

Plot everything
```{r}
dat <- data.frame(x=x_predict, t(values))
dat <- melt(dat, id="x")
# create dataframe for mean
dfmean <- data.frame(x=x_predict, y=Ef, ymin=Ef-2*sqrt(diag(Cf)), ymax=Ef+2*sqrt(diag(Cf)))
# Plot
ggplot() + 
  geom_ribbon(data=dfmean, aes(x=x_predict, ymin=ymin, ymax=ymax), fill="grey80") + 
  geom_line(data=dat, aes(x=x, y=value, color=variable)) +
  geom_line(data=dfmean, aes(x=x_predict, y=Ef), size=1) + 
  geom_point(data=obs, aes(x=x, y=y), color='red') + 
  xlab("input, x") + 
  ylab("output, f(x)")
```
















